\section{Simulating the Fully-Connected Clebsch-Gordan Tensor Product with Gaunt Tensor Products}
\label{sec:simulate_TP_Gaunt}

One way to increase the expressivity of GTP is to first reweight the inputs $\x,\y$. That is, we first create
\begin{align}
    \x'^{(\ell)}=a_{\ell}\xl{}\\
    \y'^{(\ell)}=b_{\ell}\yl{}.
\end{align}
where $a_\ell$ and $b_\ell$ are learnable weights. We then perform GTP after this reweighting and extract some output irrep(s) $\ell_3$. That is we get
\begin{align}
    (\x'\gtp \y')^{(\ell_3)}.
\end{align}

The analogous operation is fully connected CGTP. There may be multiple pairs of irreps which give a $\ell_3$ output. We can always weight and sum these to get
\begin{align}
    \sum_{\ell,\ell'}w_{\ell,\ell'}( {\xl{}} \cgtp \y^{(\ell')})^{(\ell_3)}
\end{align}
where $w_{\ell,\ell'}$ are learnable weights.

However, even if we only care about symmetric tensor products, the weighted GTP operation is strictly less expressive than fully connected CGTP.

More concretely, suppose we have nontrivial $\ell=2$ and $\ell=4$ data in our inputs. From CGTP and the selection rules we see that 
\begin{align}
    (\x^{(2)}\cgtp\y^{(2)})^{(2)}\qquad
    (\x^{(2)}\cgtp\y^{(4)})^{(2)}\\
    (\x^{(4)}\cgtp\y^{(2)})^{(2)}\qquad
    (\x^{(4)}\cgtp\y^{(4)})^{(2)}
\end{align}
are all nonzero. In particular, it is possible to create a $\ell=2$ output of 
\[(\x^{(2)}\cgtp\y^{(2)})^{(2)}+(\x^{(4)}\cgtp\y^{(4)})^{(2)}\]
with a fully connected CGTP. However, GTP instead gives a single $\ell=2$ output of form
\begin{align}
    c^2_{2,2}(\x'^{(2)}\cgtp\y'^{(2)})^{(2)}
    +c^2_{2,4}(\x'^{(2)}\cgtp\y'^{(4)})^{(2)}
    +c^2_{4,2}(\x'^{(4)}\cgtp\y'^{(2)})^{(2)}
    +c^2_{4,4}(\x'^{(4)}\cgtp\y'^{(4)})^{(2)}
\end{align}
where the $c$'s are nonzero coefficients. Note that in order to have nonzero $(\x^{(2)}\cgtp\y^{(2)})^{(2)}$ and $(\x^{(4)}\cgtp\y^{(4)})^{(2)}$ contributions, $a_2,b_2,a_4,b_4$ must all be nonzero. However, that means we must have nonzero $(\x^{(2)}\cgtp\y^{(4)})^{(2)}$ and $(\x^{(4)}\cgtp\y^{(2)})^{(2)}$ contributions. Therefore weighted GTP is not expressive enough to output  $(\x^{(2)}\cgtp\y^{(2)})^{(2)}+(\x^{(4)}\cgtp\y^{(4)})^{(2)}$, as it will necessarily mix additional terms.