\section{Discussion}
\label{sec:discussion}

The experiments in Section~\ref{sec:experiments} support a consistent
qualitative picture for the studied input: Boltz-2's reverse diffusion under
the reported configuration behaves like a sharply funneled stochastic map
toward a dominant terminal basin, while TPS+OPES still surfaces persistent
excursions to elevated ligand RMSD at non-negligible frequency along the biased
chain. Interpreting those excursions requires separating three issues: (i) what
the \emph{unbiased} path measure says about concentration; (ii) what a
two-dimensional OPES projection reveals about a \emph{margin} of the path
measure; and (iii) how much of the tail is attributable to adaptive bias updates
versus genuine secondary basins of the underlying generator.

\paragraph{Sharp funnel versus memorization.}
A single forward pass cannot distinguish a sharply legitimate ensemble from an
overly rigid memorized map. Unbiased TPS is designed to answer that distinction
at the path level: if the induced path measure were diffuse, accepted paths
would explore structurally diverse terminals at appreciable rates under the
same move set. The observed sub-angstrom RMSD spread instead indicates that,
for this input, most stochastic mass sits near one binding geometry. That
conclusion is not a sampling artifact of forward shooting: each accepted path
involves independent draws of augmentation and churn noise, so identical
terminal packing reflects the generator rather than a failure to perturb the
chain.

\paragraph{OPES projections and caveats.}
The two-dimensional OPES summaries should be read as biased-path statistics
together with the fixed-bias lemma (Lemma~\ref{lem:opes-mh}) and the adaptive-bias
remark in Section~\ref{sec:opes-theory}. Different $(\gamma, \mathrm{barrier},
\mathrm{pace})$ choices change the visited kernel mixture and therefore the
shape of reported surfaces. Secondary peaks can also arise from projection:
distinct three-dimensional configurations can map to similar $(s_1,s_2)$.
Cross-checking with diagnostic scalars logged alongside trajectories (clashes,
Ramachandran outliers, lDDT, pLDDT) remains essential before chemical
interpretation.

\paragraph{Generality.}
The same TPS construction applies to other generators cast as explicit
discrete-time Markov kernels with tractable noise priors, including many
score-based samplers and flow-matching implementations \citep{lipman2022}. The
Boltz-2-specific content is the map $T_i$, the schedule, and the augmentation
stack; Sections~\ref{sec:kernel-full}--\ref{sec:diffusion-boltz} document that
specialization pattern for reuse.

\paragraph{Relation to Boltzmann generators and action methods.}
Deep equilibrium models trained as Boltzmann generators define path measures
similar to Eq.~\eqref{eq:forward-kernel-product} and have been combined with TPS
for rare events \citep{falkner2023}. Complementary approaches interpret
diffusion as inducing dynamics whose likely paths minimize an action functional
\citep{raja2025omtps}. Our emphasis is diagnostic rather than control: we use
exact MH corrections for the given pretrained sampler rather than learning a
new path measure.

Section~\ref{sec:conclusion} collects practical takeaways and open directions.
