% ============================================================
%  Transition Path Sampling for Boltz-2 Reverse Diffusion
%  Merged theory: rigorous formulation + implementation notes
%
%  NOTE: This file is legacy standalone working notes (article class).
%  The canonical Supporting Information is the JMLR-formatted driver
%    docs/paper_tps_generative_models_SI.tex
%  with modular sources under docs/si/*.tex and references in
%    docs/si_refs.bib.  Edit those files for publication; keep this
%  document only if you still want a single-file article export.
%  The SI omits deep implementation-only notes (commit hashes, bug histories).
% ============================================================
\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{bbm}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{geometry}
\geometry{margin=2.5cm}

% ---------- theorem environments ----------
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

% ---------- notation ----------
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\slogdet}{\operatorname{slogdet}}
\newcommand{\Haar}{\mu_{\mathrm{Haar}}}
\newcommand{\xtilde}{\tilde{x}}
\newcommand{\xnoisy}{x^{\mathrm{noisy}}}
\newcommand{\xhat}{\hat{x}}
\newcommand{\that}{\hat{t}}
\newcommand{\Dtheta}{D_\theta}
\newcommand{\ftheta}{f_\theta}
\newcommand{\cskip}{c_{\mathrm{skip}}}
\newcommand{\cout}{c_{\mathrm{out}}}
\newcommand{\cin}{c_{\mathrm{in}}}
\newcommand{\cnoise}{c_{\mathrm{noise}}}
\newcommand{\sigmaD}{\sigma_{\mathrm{d}}}
\newcommand{\HAB}{H_{AB}}
\newcommand{\pacc}{p_{\mathrm{acc}}}
\newcommand{\pfwd}{p_{\mathrm{fwd}}}
\newcommand{\pbwd}{p_{\mathrm{bwd}}}
\newcommand{\Xold}{X^{\mathrm{o}}}
\newcommand{\Xnew}{X^{\mathrm{n}}}
\newcommand{\kold}{\hat{k}^{\mathrm{o}}}
\newcommand{\knew}{\hat{k}^{\mathrm{n}}}

\title{\textbf{Transition Path Sampling for Boltz-2 Reverse Diffusion}\\
\large A Rigorous Formulation with SE(3) Augmentation,
Exact Jacobians, and Implementation Notes}
\author{}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\newpage

% ============================================================
\section{Setup and Extended State Space}
\label{sec:setup}
% ============================================================

\subsection{The Boltz-2 reverse sampler as stochastic dynamics}

Boltz-2~\cite{boltz2} generates biomolecular structures by running a
reverse diffusion process: starting from Gaussian noise, the model
iteratively denoises atom coordinates until a plausible 3D structure
emerges.  We treat this reverse sampler as a
\emph{stochastic dynamical system} and apply the classical
Bolhuis--Chandler--Dellago transition path sampling
(TPS)~\cite{bolhuis2002,falkner2023} framework to it.

The diagnostic goal is to understand \emph{how} Boltz-2 brings noise
to structure: are there metastable intermediate states?  Are there
``bad'' denoising trajectories?  Which network weights are responsible
for unusual pathways?

\subsection{Extended state at diffusion step $i$}
\label{sec:extended-state}

The Boltz-2 sampler runs for $L$ discrete steps
(controlled by \texttt{num\_sampling\_steps}, default~5).  At each step
$i \in \{0,1,\ldots,L\}$ we define the \emph{extended state}
\begin{equation}
  \mathcal{S}_i = \bigl(x_i,\; R_i,\; \tau_i,\; \varepsilon_i,\; i\bigr),
\end{equation}
where:
\begin{itemize}
  \item $x_i \in \R^{M \times 3}$: atom coordinates for $M$ atoms.
  \item $R_i \in \mathrm{SO}(3)$: random rotation drawn from the Haar
        (uniform) measure at step $i$, implemented via
        \texttt{random\_quaternions} $\to$ \texttt{quaternion\_to\_matrix}.
  \item $\tau_i \in \R^{1 \times 3}$: random translation drawn from
        $\N(0, s_{\mathrm{trans}}^2 I)$ with $s_{\mathrm{trans}} = 1.0$.
  \item $\varepsilon_i \in \R^{M \times 3}$: additive noise (the
        ``churn'' noise) injected at step $i$.
  \item $i$: the step index, which serves as the ``time'' axis for TPS.
\end{itemize}
The endpoints of TPS are:
\begin{align}
  \text{State } A &: x_0 \sim \N(0,\, \sigma_0^2 I),\quad
                     \sigma_0 = \sigma_{\max},\\
  \text{State } B &: \phi(x_L) \in B,\quad
                     \phi \text{ a quality CV (pLDDT, RMSD, \ldots)}.
\end{align}

\subsection{Why include SE(3) in the extended state (Option A)}
\label{sec:optionA}

At each step Boltz-2 applies a random rigid transformation to the
current coordinates before adding noise and denoising.  There are two
possible treatments:

\begin{description}
  \item[Option~A (exact, adopted here):] Include $(R_i, \tau_i)$ in the
    extended state and store them in every snapshot.  The full path
    probability uses the complete kernel including augmentation.  No
    assumption about the network's equivariance is needed.

  \item[Option~B (approximate):] Fix $R_i = I$, $\tau_i = 0$ at all
    steps.  Valid only under \emph{exact} SE(3) equivariance of the
    network, which does not hold in practice.
\end{description}

We adopt \textbf{Option~A}.  The apparent complexity is illusory: the
Haar and translation terms \emph{cancel exactly} in acceptance ratios
(proved in \cref{sec:path-prob}), so the computational overhead is
merely storing $(R_i, \tau_i)$ per snapshot.

\paragraph{SE(3) augmentation index (implementation note).}
When \texttt{single\_backward\_step}$(x_{i+1}, i)$ runs, it samples a
fresh augmentation $(R_i, \tau_i)$ and stores it on the \emph{result}
frame $x_i$.  Thus, for a backward-generated frame $x_i$ (i.e.\ $s_0$
in the edge $(s_0, s_1)$ of the prefix), the correct augmentation to
pass to \texttt{recover\_forward\_noise} is $s_0.R$ and $s_0.\tau$, not
$s_1.R$ and $s_1.\tau$ (which belong to the next step $i+1$).  Using
the wrong frame's augmentation was a historical bug that was found and
fixed in commit \texttt{f0c6731}.

% ============================================================
\section{Full Transition Kernel with SE(3) Augmentation}
\label{sec:kernel}
% ============================================================

\subsection{The one-step map}

Each Boltz-2 reverse-diffusion step consists of three sub-steps.
The noise schedule $(\sigma_i, \gamma_i)_{i=0}^{L-1}$ and the scalars
$s$ (\texttt{step\_scale}), $\eta$ (\texttt{noise\_scale}),
$\sigmaD$ (\texttt{sigma\_data}) are fixed hyperparameters.

\paragraph{Sub-step (a): SE(3) augmentation.}
\begin{equation}
  \xtilde_i = R_i \cdot \bigl(x_i - \bar x_i\bigr) + \tau_i,
  \qquad
  R_i \sim \Haar\bigl(\mathrm{SO}(3)\bigr),\quad
  \tau_i \sim \N\bigl(0,\, s_{\mathrm{trans}}^2 I\bigr),
  \label{eq:se3-aug}
\end{equation}
where $\bar x_i = \frac{1}{M}\sum_m (x_i)_m$ is the centroid.

\paragraph{Sub-step (b): Churn noise injection.}
Define $\that_i = \sigma_{i-1}(1 + \gamma_i)$ and inject
\begin{equation}
  \xnoisy_i = \xtilde_i + \varepsilon_i,
  \qquad
  \varepsilon_i \sim \N\bigl(0,\, v_i\, I\bigr),\quad
  v_i = \eta^2\bigl(\that_i^2 - \sigma_{i-1}^2\bigr).
  \label{eq:noise-inject}
\end{equation}
When $\gamma_i = 0$, $v_i = 0$ and no noise is injected.

\paragraph{Sub-step (c): Deterministic denoising update.}
\begin{equation}
  x_{i+1} = \xnoisy_i + s\,\frac{\sigma_i - \that_i}{\that_i}
             \bigl(\xnoisy_i - \xhat_i\bigr)
  = (1 + \alpha_i)\,\xnoisy_i - \alpha_i\,\Dtheta(\xnoisy_i,\, \that_i),
  \label{eq:update-alpha}
\end{equation}
where $\alpha_i \coloneqq s(\sigma_i - \that_i)/\that_i$ and
$\xhat_i = \Dtheta(\xnoisy_i, \that_i)$ is the EDM-preconditioned
denoiser (see \cref{sec:edm-precond}).

\subsection{The transition kernel}
\label{sec:transition-kernel}

The transition kernel on the extended state is a product measure:
\begin{equation}
  p\bigl((x_{i+1}, R_i, \tau_i, \varepsilon_i) \mid x_i\bigr)
  =
  \Haar(R_i)\cdot
  \N(\tau_i;\, 0,\, s_{\mathrm{trans}}^2 I)\cdot
  \N(\varepsilon_i;\, 0,\, v_i I)\cdot
  \delta\bigl(x_{i+1} - T_i(x_i; R_i, \tau_i, \varepsilon_i)\bigr).
  \label{eq:transition-kernel}
\end{equation}

\subsection{Backward step: SE(3) inversion and fixed-point inversion}
\label{sec:backward-step}

For backward shooting, two inversions are required:

\paragraph{SE(3) inversion.}  Given $(\xtilde_i, R_i, \tau_i)$,
\begin{equation}
  x_i = R_i^\top(\xtilde_i - \tau_i) + \bar x_i.
  \label{eq:se3-inversion}
\end{equation}
Since $R_i$ is stored in the snapshot, this is exact and $O(M)$.

\paragraph{Fixed-point inversion for $\xnoisy_i$.}
The forward update satisfies
$x_{i+1} = (1+\alpha)\,\xnoisy - \alpha\,\Dtheta(\xnoisy, \that)$.
Given $x_{i+1}$, recovering $\xnoisy$ requires solving this implicit
equation.  The na\"ive fixed-point iteration
$x \leftarrow (x_{i+1} + \alpha\,D(x,\that))/(1+\alpha)$
diverges at low noise because $D(x) \approx \cskip\,x$ gives a
Jacobian with spectral radius $|\alpha\,\cskip/(1+\alpha)| > 1$ when $s > 1$.

The rescaled iteration factors out $\cskip$:
\begin{equation}
  x \;\leftarrow\; \frac{x_{i+1} + \alpha\bigl(\xhat(x,\that) - \cskip\,x\bigr)}
                        {\gamma_{\mathrm{eff}}}, \qquad
  \gamma_{\mathrm{eff}} = 1 + \alpha(1 - \cskip),
  \label{eq:fp-inversion}
\end{equation}
whose Jacobian is $\alpha(D'(x) - \cskip)/\gamma_{\mathrm{eff}}$,
which is bounded and small at all noise levels.  Convergence in 4
iterations is sufficient in practice.

\paragraph{Center-mean consistency (implementation note).}
The backward proposal in \texttt{prefix\_backward\_proposal\_log\_prob\_tensor}
uses the per-atom mean of the \emph{lower}-frame coordinates $x_j$ as the
centering reference (not a stored \texttt{center\_mean\_before\_step}).
This is consistent with how \texttt{single\_backward\_step} centers before
augmentation.  An earlier version used the stored value, which was a minor
inconsistency fixed in commit \texttt{f0c6731}.

% ============================================================
\section{The Underlying Stochastic Differential Equation}
\label{sec:sde}
% ============================================================

\subsection{The VESDE forward process}

Following Song et al.~\cite{song2021sde}, the Boltz-2 forward process
is a Variance-Exploding SDE (VESDE):
\begin{equation}
  dx = \sqrt{\frac{d\sigma^2(t)}{dt}}\,dW_t,
  \qquad
  p(x_t \mid x_0) = \N\bigl(x_t;\, x_0,\, \sigma(t)^2 I\bigr).
  \label{eq:vesde-forward}
\end{equation}

\subsection{Reverse probability flow ODE and score--denoiser identity}

The reverse probability flow ODE in $\sigma$-space:
\begin{equation}
  \frac{dx}{d\sigma} = \frac{x - \Dtheta(x, \sigma)}{\sigma},
  \label{eq:reverse-ode}
\end{equation}
where Tweedie's formula gives:
\begin{equation}
  \nabla_x \log p_\sigma(x)
  = \frac{\Dtheta(x, \sigma) - x}{\sigma^2}.
  \label{eq:score-denoiser}
\end{equation}

\subsection{EDM preconditioning}
\label{sec:edm-precond}

Karras et al.~\cite{karras2022edm} introduce preconditioning:
\begin{equation}
  \Dtheta(x, \sigma)
  = \cskip(\sigma)\,x
  + \cout(\sigma)\,\ftheta\!\bigl(\cin(\sigma)\,x,\; \cnoise(\sigma)\bigr),
  \label{eq:edm-precond}
\end{equation}
with the scalars (\texttt{diffusionv2.py}, lines 239--249):
\begin{align}
  \cskip(\sigma) &= \frac{\sigmaD^2}{\sigma^2 + \sigmaD^2},
  &
  \cout(\sigma)  &= \frac{\sigma\,\sigmaD}{\sqrt{\sigma^2 + \sigmaD^2}},
  \\
  \cin(\sigma)   &= \frac{1}{\sqrt{\sigma^2 + \sigmaD^2}},
  &
  \cnoise(\sigma)&= \tfrac{1}{4}\ln\!\bigl(\sigma/\sigmaD\bigr).
\end{align}

\subsection{Boltz-2 sampler as Karras Algorithm~2}
\label{sec:karras-alg2}

The Boltz-2 reverse sampler implements Karras Algorithm~2 (stochastic
sampler with churn):
\begin{enumerate}
  \item \textbf{Churn}: raise the noise level from $\sigma_{i-1}$ to
        $\that_i = \sigma_{i-1}(1+\gamma_i)$ by adding
        $\varepsilon_i \sim \N(0, v_i I)$.
  \item \textbf{ODE step}: take a first-order Euler step
        $x_{i+1} = \xnoisy_i + (\sigma_i - \that_i)\,\frac{\xnoisy_i - \Dtheta(\xnoisy_i, \that_i)}{\that_i}$.
\end{enumerate}

\begin{remark}[The churn is a finite Langevin correction, not EM]
  For $\gamma_i = 0.8$ (Boltz-2 default), the churn noise variance is
  $\eta^2\gamma_i(2+\gamma_i)\sigma_{i-1}^2$, a \emph{finite} amount
  independent of the step size $|\sigma_i - \sigma_{i-1}|$.  This is
  \emph{not} an Euler--Maruyama discretisation of a fixed SDE.  As
  $\gamma_i \to 0$, the stochasticity vanishes and the sampler becomes
  the deterministic probability flow ODE.
\end{remark}

\subsection{Discrete-time Markov chain: the rigorous framework}

For TPS purposes, the rigorous foundation is the
\emph{continuous-state, discrete-time, inhomogeneous Markov chain}
$\{p_i(x_{i+1} \mid x_i)\}_{i=0}^{L-1}$.
Mora, Walczak, and Zamponi~\cite{mora2012tps} showed that TPS is
well-defined for any such process provided transition kernel ratios
can be computed.  Both conditions are satisfied here.

% ============================================================
\section{Path Probability and the Jacobian}
\label{sec:path-prob}
% ============================================================

\subsection{Log path probability}

There are two useful conventions.  In the extended random-variable
parameterization of \cref{eq:transition-kernel}, the sampled variables
$(R_i,\tau_i,\varepsilon_i)$ are part of the state and no
change-of-variables Jacobian appears:

\begin{equation}
  \log P_{\rm ext}(\mathcal{X}_{\rm ext})
  = \log\rho(x_0)
  + \sum_{i=0}^{L-1}
    \Bigl[
      \underbrace{\log\Haar(R_i)}_{\text{constant}}
      + \underbrace{\log\N(\tau_i;\, 0,\, s_{\mathrm{trans}}^2 I)}_{\text{Gaussian}}
      + \underbrace{\log\N(\varepsilon_i;\, 0,\, v_i I)}_{\text{Gaussian}}
    \Bigr],
  \label{eq:log-path-prob}
\end{equation}

If one instead reports a coordinate-density diagnostic, the noise-to-coordinate
map contributes $-\log|\det J_i|$, where
$J_i = \partial x_{i+1}/\partial\varepsilon_i$.  The production code's
``trajectory_log_path_prob`` diagnostic includes the initial prior, the
$\varepsilon_i$ priors, and an optional scalar schedule-only approximation to
this Jacobian.  It omits the Haar factors and the nonconstant $\tau_i$ priors,
which cancel in the implemented MH ratios because the same translation prior
appears in proposal and target factors.

\subsection{The exact Jacobian $J_i$}
\label{sec:jacobian}

From \cref{eq:update-alpha} with $\xnoisy_i = \xtilde_i + \varepsilon_i$:
\begin{equation}
  \frac{\partial x_{i+1}}{\partial\varepsilon_i}
  = (1+\alpha_i)I - \alpha_i\,\frac{\partial\Dtheta}{\partial x}\bigg|_{\xnoisy_i}.
  \label{eq:jacobian-raw}
\end{equation}
Expanding via \cref{eq:edm-precond}:
\begin{equation}
  \boxed{
  J_i
  = \beta_i\,I + \mu_i\,\nabla_z\ftheta\bigg|_{z=\cin(\that_i)\,\xnoisy_i},
  }
  \label{eq:jacobian-exact}
\end{equation}
with the scalars
\begin{align}
  \beta_i &= 1 + \alpha_i\bigl(1 - \cskip(\that_i)\bigr),
  \label{eq:beta}\\
  \mu_i   &= -\alpha_i\,\cout(\that_i)\,\cin(\that_i).
  \label{eq:mu}
\end{align}

\begin{remark}
  The claim that $J_i$ is a scalar multiple of the identity is
  \emph{incorrect}.  The Jacobian $\nabla_z\ftheta$ is a full
  $3M\times 3M$ matrix and is not proportional to $I$ in general.
  The scalar approximation $\log|\det J_i| \approx 3M\log|1+\alpha_i|$
  (setting $\nabla_z\ftheta = 0$) is nonetheless \emph{exact for
  MCMC acceptance ratios} at fixed schedule because $\alpha_i$ depends
  only on the step index; see \cref{prop:oneway-cancel}.
\end{remark}

\subsection{Log-determinant decomposition}
\label{sec:logdet}

\begin{equation}
  \log|\det J_i|
  = 3M\log|\beta_i|
  + \log\bigl|\det\bigl(I + c_i\,\nabla_z\ftheta\bigr)\bigr|,
  \qquad
  c_i = \frac{\mu_i}{\beta_i}.
  \label{eq:logdet-decomp}
\end{equation}

\subsection{Exact computation via chunked forward-mode AD}
\label{sec:chunked-ad}

The Jacobian $\nabla_z\ftheta \in \R^{3M\times 3M}$ is assembled
column-by-column using JVPs via \texttt{torch.func.jvp} wrapped in
\texttt{torch.func.vmap} with chunk size $C$.  The log-determinant is
then computed by \texttt{torch.linalg.slogdet}.

\begin{table}[h]
\centering
\caption{Cost of one exact Jacobian log-determinant evaluation (A100,
  $T_{\mathrm{fwd}}\approx 50\,\mathrm{ms}$, $C=64$).}
\label{tab:jacobian-cost}
\begin{tabular}{rrrrc}
\toprule
$M$ (atoms) & $3M$ & Chunks & $\approx$ time/step & Feasible? \\
\midrule
  50 &  150 &  3 & 0.15\,s & \checkmark \\
 200 &  600 & 10 & 0.50\,s & \checkmark \\
 500 & 1500 & 24 & 1.2\,s  & \checkmark \\
1000 & 3000 & 47 & 2.4\,s  & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cancellation of Jacobians in acceptance ratios}
\label{sec:jacobian-cancellation}

\begin{proposition}[One-way forward shooting: scalar Jacobians cancel]
\label{prop:oneway-cancel}
For one-way forward shooting from shooting point $k$, the scalar
schedule-dependent Jacobian approximation cancels exactly in the
Metropolis acceptance ratio.
\end{proposition}

\begin{proof}
In one-way forward shooting, the new forward segment $(k\to L)$ is
generated by drawing fresh $(R_i^{\mathrm{n}}, \tau_i^{\mathrm{n}},
\varepsilon_i^{\mathrm{n}})$ from their priors for $i=k,\ldots,L-1$.
The generation probability for the new suffix is
$G(\Xnew \mid \Xold, k) = \prod_{i=k}^{L-1}
\Haar(R_i^{\mathrm{n}})\N(\tau_i^{\mathrm{n}})\N(\varepsilon_i^{\mathrm{n}})$.
The path probability ratio for the new suffix is
$\prod_{i=k}^{L-1} \N(\varepsilon_i^{\mathrm{n}}) / |\det J_i^{\mathrm{n}}|$
divided by the same for the old.
The Metropolis acceptance is $P(\Xnew)/P(\Xold) \times G(\Xold)/G(\Xnew)$.
The $\Haar$, $\N(\tau)$, and $\N(\varepsilon)$ terms from $G$ cancel
the corresponding terms from $P$ in both numerator and denominator.
Under the scalar approximation, the Jacobian terms
$|\det J_i^{\mathrm{n}}|$ appear in the $P$ ratio but not in $G$;
however, they cancel because $\alpha_i$ depends only on the schedule
step $i$ (not on $\varepsilon_i$ or $x_i$), so
$|\det J_i^{\mathrm{n}}| = |\det J_i^{\mathrm{o}}|$ for each
$i \geq k$.  Exact Jacobians need not be path-independent and therefore
must be treated as diagnostics unless the full proposal ratio is derived
with those state-dependent determinants.
The remaining acceptance ratio is $\HAB(\Xnew)$.
\end{proof}

\begin{proposition}[Two-way shooting: Jacobians required for backward segment]
\label{prop:twoway-jacobian}
For two-way shooting, Jacobian log-determinants appear in the
backward segment ($0$ to $k-1$) of the acceptance ratio, not in the
forward segment.
\end{proposition}

\begin{proof}
The forward segment ($k\to L$) of two-way shooting uses the same
cancellation as \cref{prop:oneway-cancel}.  For the backward segment
($k\to 0$), the new path is generated by the re-noising kernel
$\pbwd$, whereas the path probability uses the forward kernel $\pfwd$.
These kernels are different: $\pbwd$ simply adds Gaussian noise, while
$\pfwd$ contains the deterministic Boltz-2 update with Jacobian.
The acceptance ratio for the backward segment is (see \cref{eq:twoway-acceptance-full}).
\end{proof}

% ============================================================
\section{TPS in the Extended Ensemble}
\label{sec:extended-ensemble}
% ============================================================

\subsection{Fixed-length path ensemble}

Following Falkner et al.~\cite{falkner2023}, define the extended
ensemble measure
\begin{equation}
  \pi(\mathcal{X}, \hat k)
  \propto
  P_{AB}(\mathcal{X})\,P(\mathcal{X})\,p(\hat k \mid \mathcal{X}),
\end{equation}
where $P_{AB}(\mathcal{X}) = \mathbbm{1}[x_0 \in A,\, x_L \in B]$ and
$p(\hat k \mid \mathcal{X})$ is the shooting point distribution
(e.g.\ uniform over $\{1,\ldots,L-1\}$).

\subsection{One-way forward shooting: simplified acceptance}

\begin{equation}
  \pacc = \HAB(\Xnew).
  \label{eq:oneway-acceptance}
\end{equation}

\subsection{Two-way shooting: full acceptance ratio}

\begin{equation}
  \pacc = \HAB(\Xnew)\cdot
  \min\!\Biggl[1,\;
    \frac{\rho(x_0^{\mathrm{n}})}{\rho(x_0^{\mathrm{o}})}
    \prod_{i=0}^{k-1}
    \frac{\pfwd(x_{i+1}^{\mathrm{n}} \mid x_i^{\mathrm{n}})\,
          \pbwd(x_i^{\mathrm{o}} \mid x_{i+1}^{\mathrm{o}})}
         {\pfwd(x_{i+1}^{\mathrm{o}} \mid x_i^{\mathrm{o}})\,
          \pbwd(x_i^{\mathrm{n}} \mid x_{i+1}^{\mathrm{n}})}
  \Biggr],
  \label{eq:twoway-acceptance-full}
\end{equation}
where
\begin{align}
  \log\pfwd(x_{i+1}\mid x_i)
  &= -\frac{\|\varepsilon_i^*\|^2}{2v_i}
     - \frac{3M}{2}\log(2\pi v_i)
     - \log|\det J_i|,
  \label{eq:pfwd-logprob}\\
  \log\pbwd(x_i \mid x_{i+1})
  &= -\frac{\|\varepsilon_i^{\mathrm{bwd}}\|^2}{2v_i}
     - \frac{3M}{2}\log\bigl(2\pi v_i\bigr).
  \label{eq:pbwd-logprob}
\end{align}
In log form (as implemented in \texttt{backward\_shooting\_metropolis\_bias}):
\begin{equation}
  \log r
  = (\mathrm{lf\_new} - \mathrm{lf\_old})
  + (\mathrm{lb\_old} - \mathrm{lb\_new})
  + (\log\rho(x_0^{\mathrm{n}}) - \log\rho(x_0^{\mathrm{o}})).
  \label{eq:logr-implementation}
\end{equation}

\paragraph{Common sign-error pitfall (historical bug).}
Reversing the signs of the $\mathrm{lf}$ and $\mathrm{lb}$ terms
(computing $\mathrm{lf\_old} - \mathrm{lf\_new} + \mathrm{lb\_new} -
\mathrm{lb\_old}$) gives $\log(1/r)$ instead of $\log r$.  When $r < 1$
(move should be penalised), the inverted formula yields
$\exp(-\log r) > 1$, which clamps to 1 and always accepts.  When $r > 1$
(move should freely accept), the inverted formula gives $< 1$ and
incorrectly rejects.  Both effects together completely invert the
sampling, producing unphysical (``exploded'') terminal structures.  This
bug was identified and fixed in commit \texttt{f0c6731}.

% ============================================================
\section{Shooting Moves}
\label{sec:shooting}
% ============================================================

\subsection{Forward shooting}
\label{sec:forward-shooting}

\begin{enumerate}
  \item Select shooting index $k$ uniformly from $\{1,\ldots,L-1\}$.
  \item \emph{(Optional)} Perturb the shooting point:
        $x_k \leftarrow x_k + \delta$, $\delta \sim \N(0,
        \sigma_{\mathrm{pert}}^2 I)$.
  \item Keep $(x_0,R_0,\tau_0,\varepsilon_0),\ldots,x_k$ from the old path.
  \item Draw fresh $(R_k, \tau_k, \varepsilon_k), \ldots,
        (R_{L-1}, \tau_{L-1}, \varepsilon_{L-1})$ from their priors.
  \item Propagate: $x_{i+1} = T_i(x_i; R_i, \tau_i, \varepsilon_i)$
        for $i = k, \ldots, L-1$.
  \item Accept if $\HAB(\Xnew) = 1$; no Jacobian needed
        (\cref{prop:oneway-cancel}).
\end{enumerate}

\subsection{Backward shooting (two-way)}
\label{sec:backward-shooting}

\begin{enumerate}
  \item Select shooting index $k$ uniformly from $\{1,\ldots,L-1\}$.
  \item Keep $x_k, (R_k, \tau_k, \varepsilon_k), \ldots, x_L$ from
        the old path.
  \item Generate the new backward segment by re-noising:
        for $i = k-1, \ldots, 0$, apply
        \texttt{single\_backward\_step}$(x_{i+1}, i)$ which:
        (a) recovers $\xnoisy_i$ via the rescaled fixed-point iteration
        (\cref{eq:fp-inversion}),
        (b) samples fresh $(\varepsilon_i', R_i', \tau_i')$,
        (c) stores $(R_i', \tau_i')$ on the result frame $x_i'$.
  \item Run the forward segment $k \to L$ as in forward shooting.
  \item Compute the MH acceptance ratio via
        \cref{eq:twoway-acceptance-full}:
        \begin{itemize}
          \item For the scalar (approximate) Jacobian: use
            \texttt{log\_det\_jacobian\_step}($\alpha_i$, $M$).
            These terms cancel in the ratio (\cref{prop:oneway-cancel}),
            so including or excluding them has no effect.
          \item For the exact Jacobian (diagnostic mode only): compute
            \texttt{compute\_log\_det\_jacobian\_exact} from
            \texttt{path\_probability.py}; do not use these
            state-dependent determinants in
            \texttt{backward\_shooting\_metropolis\_bias} without a
            separately derived proposal ratio.
        \end{itemize}
  \item Accept with probability $\pacc$ from \cref{eq:logr-implementation}.
\end{enumerate}

\begin{remark}[Cost comparison]
  Forward shooting costs $(L-k)$ network calls with no Jacobian.
  Backward shooting costs $k$ calls for the re-noising plus
  optionally $k \cdot \lceil 3M/C\rceil$ calls for exact Jacobians.
  For $L=5$, $k=2$, $M=200$, $C=64$: exact Jacobian overhead is
  $\approx 20$ extra forward passes ($\approx 1.0\,\mathrm{s}$ on an A100).
\end{remark}

% ============================================================
\section{Ergodicity of the Full Move Set}
\label{sec:ergodicity}
% ============================================================

Define the \emph{reactive path ensemble} $\mathcal{R}$ as fixed-length
paths $X = (x_0,\ldots,x_L)$ with $x_0 \in A$ and $x_L \in B$,
weighted by $\pi(X) \propto P(\mathcal{X})\,\mathbf{1}[X \in \mathcal{R}]$.

\paragraph{Forward-only chain.}
Ergodic over sub-ensembles with fixed $x_0$, but not over $\mathcal{R}$:
paths with different starting noise configurations $x_0$ are never reached.

\paragraph{Backward-only chain.}
Not ergodic over $\mathcal{R}$ because $x_L$ is never resampled.

\paragraph{Forward + backward chain (two-way shooting).}
Backward moves can change $x_0$ (choosing $k=1$) while forward moves
can change $x_L$ (choosing $k=L-1$).

\begin{theorem}[Ergodicity of two-way shooting]
\label{thm:ergodicity}
Suppose the reactive path space is connected under the move set.
Then the Markov chain on $\mathcal{R}$ defined by forward + backward
shooting with the acceptance rules in \cref{eq:twoway-acceptance-full}
is ergodic with stationary distribution $\pi$.
\end{theorem}

\begin{proof}[Proof sketch]
Standard Metropolis--Hastings theory~\cite{bolhuis2002}: each move
satisfies detailed balance with respect to $\pi$ by construction.
Irreducibility follows from the assumed connectivity of path space.
\end{proof}

\paragraph{Global reshuffle move.}
For additional robustness, draw $x_0' \sim \rho(\cdot)$ and all
$(\varepsilon_i', R_i', \tau_i')$ fresh, then forward-integrate.
The acceptance ratio is
\begin{equation}
  \frac{\pi(X')}{\pi(X)} \cdot
  \frac{q_{\mathrm{reshuffle}}(X \to X)}{q_{\mathrm{reshuffle}}(X \to X')}
  = \frac{\rho(x_0') \prod_i \pi_i(\xi_i')}
         {\rho(x_0) \prod_i \pi_i(\xi_i)}
    \cdot \frac{\rho(x_0) \prod_i \pi_i(\xi_i)}
               {\rho(x_0') \prod_i \pi_i(\xi_i')}
  = 1,
\end{equation}
so the reshuffle always accepts (when the trial path is reactive).
It reaches any path in a single step, guaranteeing irreducibility.

\paragraph{Complete move scheme.}
\begin{itemize}
  \item 45\% forward shooting (uniform selector; always accepts);
  \item 45\% backward shooting (uniform selector; explicit MH via
    \cref{eq:logr-implementation});
  \item 10\% global reshuffle (always accepts).
\end{itemize}
This is ergodic on $\mathcal{R}$ without connectivity assumptions.

\paragraph{NaN guard (implementation note).}
The Boltz-2 model can occasionally produce NaN outputs (e.g.\ when
\texttt{weighted\_rigid\_align} receives non-finite inputs).  Multiple
defence layers are required:
\begin{enumerate}
  \item \texttt{weighted\_rigid\_align} returns early on non-finite input.
  \item \texttt{\_nan\_fallback()} in \texttt{gpu\_core.py} replaces
    non-finite values with the previous state.
  \item \texttt{trajectory\_has\_finite\_coords()} sets $\mathrm{bias}=0$
    for NaN trials, preventing acceptance.
  \item An \texttt{isfinite} check in OPS's Metropolis criterion ensures
    NaN probabilities are never accepted (NaN $> u$ evaluates to False,
    which would incorrectly accept all NaN trials without this guard).
\end{enumerate}
All four layers are implemented in commit \texttt{f0c6731}.

% ============================================================
\section{Collective Variables for Diagnostics}
\label{sec:cvs}
% ============================================================

The following collective variables (CVs) are available in
\texttt{collective\_variables.py}.  They are evaluated per MC step on the
current accepted path and logged to \texttt{cv\_values.json} alongside the
OPES bias CV.  \emph{Diagnostic CVs do not feed into the OPES bias.}

\subsection*{Bias CV (sampling)}
\begin{description}
  \item[OpenMM Cα-RMSD to local minimum (\textbf{sole bias CV})]
    Each last frame of the TPS trajectory is energy-minimised with
    AMBER14/GBn2 (OpenMM, L-BFGS).  The Kabsch-aligned Cα-RMSD between
    the raw generated structure and its own local minimum is used as the
    OPES collective variable.  Structures near a stable fold have low RMSD;
    hallucinated or disordered structures have high RMSD.
\end{description}

\subsection*{Geometric diagnostic CVs}
\begin{description}
  \item[Contact order]
    Fraction of long-range Cα pairs (sequence separation $\geq 6$,
    threshold $8\,\text{\AA}$):
    \[
      \mathrm{CO}(x) = \frac{|\{(i,j): |i-j|\geq 6,\; d_{ij}<8\,\text{\AA}\}|}
                            {|\{(i,j): |i-j|\geq 6\}|}
    \]
    Detects \textbf{mode collapse} (near-zero variance across ensemble implies
    constant CO) and fold topology changes.  Implemented as
    \texttt{contact\_order} in \texttt{collective\_variables.py}.

  \item[Steric clash count]
    Number of Cα pairs with $|i-j|\geq 3$ and $d_{ij}<3.0\,\text{\AA}$
    (each pair counted once):
    \[
      N_{\mathrm{clash}}(x) = |\{(i,j): |i-j|\geq 3,\; d_{ij}<3.0\,\text{\AA},\; i<j\}|
    \]
    Detects \textbf{unphysical packing}, collapsed structures, and raw Boltz
    output artefacts before OpenMM minimisation.  Implemented as
    \texttt{clash\_count}.

  \item[lDDT to reference]
    Alignment-free local distance difference test \cite{Mariani2013}:
    for each pair within $15\,\text{\AA}$ in the reference,
    compute fraction of $|d^{\mathrm{query}}_{ij}-d^{\mathrm{ref}}_{ij}|$
    below thresholds $\{0.5, 1.0, 2.0, 4.0\}\,\text{\AA}$, average:
    \[
      \mathrm{lDDT}(x, x_{\mathrm{ref}}) =
        \frac{1}{4}\sum_{t\in\{0.5,1,2,4\}}
        \frac{|\{(i,j): d^{\mathrm{ref}}_{ij}<15,\;
              |d^{\mathrm{query}}_{ij}-d^{\mathrm{ref}}_{ij}|<t\}|}
             {|\{(i,j): d^{\mathrm{ref}}_{ij}<15\}|}
    \]
    Detects \textbf{memorisation} ($\mathrm{lDDT}\approx 1$ implies
    near-exact reproduction of the reference, likely a training set structure)
    and local structural errors masked by good global RMSD.  Implemented as
    \texttt{lddt\_to\_reference}.

  \item[Ramachandran outlier fraction]
    Fraction of residues with $(\phi,\psi)$ backbone dihedral angles outside
    the simplified ``allowed'' Ramachandran regions
    (Lovell et~al.\ 2003 contour approximation):
    \[
      f_{\mathrm{Rama}}(x) = \frac{N_{\mathrm{outlier}}}{N_{\mathrm{classified}}}
    \]
    Detects \textbf{unphysical backbone geometry}, hallucinated secondary
    structure, and model shortcuts that satisfy RMSD but violate local
    chemistry.  Implemented as \texttt{ramachandran\_outlier\_fraction}.

  \item[Radius of gyration]
    $R_g(x) = \sqrt{\frac{1}{M}\sum_m \|x^m - \bar x\|^2}$
    (Ångström, no mass weighting).
    Detects global compaction / elongation changes along TPS trajectories.
    Implemented as \texttt{radius\_of\_gyration}.
\end{description}

\subsection*{Model self-confidence CV}
\begin{description}
  \item[Boltz pLDDT (model self-confidence)]
    Mean per-residue pLDDT (0--100 scale) from the Boltz-2 confidence head
    applied to the final denoised prediction.  Tracks how ``confident'' the
    model is in its own output.  High pLDDT with large OpenMM RMSD indicates
    \textbf{miscalibration} or overconfident hallucination.  Wired via
    \texttt{make\_boltz\_plddt\_predictor} in \texttt{collective\_variables.py};
    requires a loaded \texttt{Boltz2} model instance.

  \item[Step index $i$] Serves as the simplest reaction coordinate.
    The TPS committor $q(x_i) \approx i/L$ is a useful reference.

  \item[Noise level $\sigma_i$] Sigma-based state volumes for state A
    ($\sigma_i \geq \sigma_A$) and state B ($\sigma_i \leq \sigma_B$)
    as used in \texttt{states.py}.
\end{description}

\subsection*{CVs not recommended at this stage}
\begin{itemize}
  \item \textbf{TM-score to training set}: requires a database of training
    structures and TM-align; better as post-hoc analysis on saved checkpoints.
  \item \textbf{Attention entropy}: requires hooking into Boltz internals at
    every denoising step; high implementation cost for uncertain diagnostic value.
  \item \textbf{Pairwise diversity} (mean pairwise RMSD, cluster count):
    ensemble-level statistics, not per-step CVs; compute post-hoc from
    saved checkpoint NPZ files.
\end{itemize}

% ============================================================
\section*{References}
% ============================================================

\begin{thebibliography}{99}

\bibitem{boltz2}
S.~Passaro, G.~Corso, J.~Wohlwend, M.~Reveiz, S.~Thaler, V.R.~Somnath,
N.~Getz, T.~Portnoi, J.~Roy, H.~Stark, D.~Kwabi-Addo, D.~Beaini,
T.~Jaakkola, R.~Barzilay.
\newblock Boltz-2: Towards Accurate and Efficient Binding Affinity Prediction.
\newblock \textit{bioRxiv}, 2025.06.14.659707, 2025.

\bibitem{boltz1}
J.~Wohlwend, G.~Corso, S.~Passaro, M.~Reveiz, K.~Leidal, W.~Swiderski,
T.~Portnoi, I.~Chinn, J.~Silterra, T.~Jaakkola, R.~Barzilay.
\newblock Boltz-1: Democratizing Biomolecular Interaction Modeling.
\newblock \textit{bioRxiv}, 2024.11.19.624167, 2024.

\bibitem{bolhuis2002}
P.G.~Bolhuis, D.~Chandler, C.~Dellago, P.L.~Geissler.
\newblock Transition path sampling: Throwing ropes over rough mountain passes,
  in the dark.
\newblock \textit{Annu.\ Rev.\ Phys.\ Chem.}, 53:291--318, 2002.

\bibitem{falkner2023}
S.~Falkner, A.~Coretti, S.~Romano, P.~Geissler, C.~Dellago.
\newblock Revisiting Shooting Point Monte Carlo Methods for Transition Path
  Sampling.
\newblock \textit{arXiv:2408.03054}, 2023.

\bibitem{mora2012tps}
T.~Mora, A.M.~Walczak, F.~Zamponi.
\newblock Transition path sampling algorithm for discrete many-body systems.
\newblock \textit{Phys.\ Rev.\ E}, 85:036710, 2012.
\newblock \textit{arXiv:1202.0622}.

\bibitem{song2021sde}
Y.~Song, J.~Sohl-Dickstein, D.P.~Kingma, A.~Kumar, S.~Ermon, B.~Poole.
\newblock Score-Based Generative Modeling through Stochastic Differential
  Equations.
\newblock \textit{ICLR}, 2021.
\newblock \textit{arXiv:2011.13456}.

\bibitem{karras2022edm}
T.~Karras, M.~Aittala, T.~Aila, S.~Laine.
\newblock Elucidating the Design Space of Diffusion-Based Generative Models.
\newblock \textit{NeurIPS}, 2022.
\newblock \textit{arXiv:2206.00364}.

\bibitem{sb2025}
V.~De~Bortoli, V.~Durmus, M.~Hutchinson, A.~Durmus.
\newblock Foundations of Schr\"odinger Bridges for Generative Modeling.
\newblock \textit{arXiv:2603.18992}, 2025.

\bibitem{buijsman2020}
P.~Buijsman and P.~G.~Bolhuis.
\newblock Transition path sampling for non-equilibrium dynamics without
  predefined reaction coordinates.
\newblock \textit{J.~Chem.\ Phys.} \textbf{152}, 044107 (2020).

\bibitem{singlimmer2025}
A.~N.~Singh and D.~T.~Limmer.
\newblock Reactive path ensembles within nonequilibrium steady states.
\newblock \textit{arXiv:2501.19233} (2025).

\bibitem{brotzakis2016}
Z.~F.~Brotzakis and P.~G.~Bolhuis.
\newblock A one-way shooting algorithm for transition path sampling of
  asymmetric barriers.
\newblock \textit{J.~Chem.\ Phys.} \textbf{145}, 164112 (2016).

\end{thebibliography}

\end{document}
